\section{Methodology}
\label{sec:methodology}

\para{Research question.} Our main question is whether LayerNorm contributes to collapse toward stereotyped creative outputs by compressing the model's usable concept geometry. We evaluate three hypotheses. H1: an LN-free GPT-2 variant will generate more diverse responses under matched sampling. H2: the LN-free variant will show less collapsed hidden-state geometry, measured as lower anisotropy and higher effective rank. H3: if LayerNorm contributes to associative collapse, same-group vs.\ different-group clue separability should improve on \ocw.

\para{Models.} We compare the Hugging Face release \texttt{gpt2} against \texttt{schaeff/gpt2-small\_LNFree300}. We treat vanilla \gpttwo as the architectural baseline and \lnfreegpt as the intervention model. Both models are evaluated without additional fine-tuning or prompt engineering.

\para{Datasets.} We use two complementary benchmarks. For creative generation, we select 20 prompts from the creativity subset of \noveltybench, stored in \texttt{datasets/novelty\_bench/curated.jsonl}. For association geometry, we use 62 walls from the validation split of \ocw, stored in \texttt{datasets/ocw/dataset/validation.json}. The first benchmark measures output-side diversity, while the second probes representational structure on a distractor-heavy association task.

\para{Generation protocol.} For each creativity prompt and each model, we generate 12 responses with fixed sampling: temperature $=0.9$, top-$p=0.95$, \texttt{max\_new\_tokens}$=80$, and seed $=42$. This design follows prior diversity work in evaluating repeated generations instead of a single sample \citep{zhang2025noveltybench,haase2025creativity}. For prompt-following sanity checks, we also score sentence-count adherence on the subset of prompts that explicitly requested a sentence count.

\para{Creative-output metrics.} We report \texttt{distinct\_1} and \texttt{distinct\_2} for lexical diversity, mean pairwise sentence-embedding cosine distance for semantic spread, mean pairwise bigram Jaccard overlap as a self-similarity proxy, and mean repetition rate as a degeneration proxy. Higher embedding dispersion indicates broader semantic exploration, while higher bigram Jaccard indicates more overlap and therefore worse diversity.

\para{Association-geometry metrics.} For each wall, we encode clue strings with the model's final hidden layer and compute four metrics. \texttt{association\_auc} is the AUROC for classifying same-group vs.\ different-group clue pairs from cosine similarity. \texttt{similarity\_gap} is mean within-group minus between-group similarity. \texttt{anisotropy} is mean pairwise cosine among clue embeddings. \texttt{effective\_rank} is a participation-ratio estimate of covariance dimensionality. Under our hypothesis, lower anisotropy and higher effective rank indicate less representational collapse.

\para{Statistical testing.} The unit of analysis is the prompt for generation metrics and the wall for geometry metrics. For each paired comparison, we first run Shapiro--Wilk on the paired differences. If normality is not rejected, we use a paired $t$-test; otherwise we use the Wilcoxon signed-rank test. All tests are two-sided. We then apply Benjamini--Hochberg false discovery rate correction within each metric family. We report effect sizes from the saved analysis pipeline and include 95\% bootstrap confidence intervals for model means.

\para{Implementation details.} The study runs with Python 3.12.8, \texttt{torch 2.11.0+cu130}, \texttt{transformers 5.7.0}, \texttt{sentence-transformers 5.4.1}, \texttt{scikit-learn 1.8.0}, \texttt{scipy 1.17.1}, \texttt{pandas 3.0.2}, \texttt{matplotlib 3.10.9}, and \texttt{seaborn 0.13.2}. The workspace exposed four NVIDIA RTX A6000 GPUs with 49\,GB memory each, but CUDA was unavailable because the installed Torch build expected a newer driver than the container provided. As a result, all experiments ran on CPU and completed in 262.65 seconds. The main script is \texttt{src/run\_layernorm\_study.py}, with raw generations and metadata saved under \texttt{results/layernorm\_study/}.
